Cultural Transmission and Technological Transitions during the Late Paleolithic in Korea

### 2022-11-28 analysis/paper/paper.qmd
editing the results section

abstract: |
  The onset of the Late Paleolithic period in Korea, represented by the appearance of stemmed points and blades, was a key event in the dispersal of modern humans in Northeast Asia. Previous studies have mainly focussed on possible origin locations of these new technologies. The specific cultural processes of the appearance stemmed points and blades has rarely been considered. We investigate the cultural processes by applying a cultural transmission framework to investigate the social contexts of the emergence of these new technologies. Our main question is: what was the dominant mode of cultural transmission during the time of technological innovation in the Korean Late Paleolithic? Following Bettinger and Eerkens [-@bettinger1999point], we evaluated models of guided variation and  indirect bias. To evaluate these models and understand the transmission processes, we computed coefficients of variation (CV), and Principal Component Analysis (PCA). We found that information about the new technology was likely transmitted via selective combinations of guided variation and indirect bias. Some attributes including length and width were transmitted through indirect bias, while other attributes appear to have been more associated to guided variation. Our results suggest that individuals or groups learned some part of the technology of stemmed points by copying a model from another another individual or group and then modified the shape according to their specific circumstances.

@@

The application of evolutionary theory to archaeological research has been productive for the investigation of technological transitions and related human behaviors in the remote past [@dunnell1980evolutionary; @lipo1997population; @bettinger1997rediscovering; @bettinger1999point; @mesoudi2008cultural]. For example, archaeologists have used evolutionary theory and methods to study human behavioral ecology, cultural transmission, artifact phylogenetics, and niche construction in the past [@riede2010isn; @o2017dual; @garvey2018current]. In this paper we used cultural transmission theory to investigate technological transitions during the Korean Late Paleolithic. This technological innovation is not just the appearance of new tools such as stemmed points and blades, but if the appearance of these new technologies represents the first arrival of modern humans in this region, this transition may also represent a key event related to modern human dispersal throughout East Asia [@seong2009emergence]. Previous studies on the origin of stemmed points have mainly focused on their possible origin locations, connecting Korea with global patterns of modern human dispersal. However, questions about the specific processes of this technological change in Korea remain largely unanswered [@seong2008tanged; @bae2010origin; @bae2017origin]. In this research, we apply a cultural transmission framework to investigate the social contexts of the emergence of new technologies in the Korean Late Paleolithic. Our main question is: what was the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic? Additionally we ask: Do the modes of cultural transmission vary over time and space? We consider three possible modes of cultural transmission: guided variation (trial and error), indirect bias (copying others), or a combination of the two. Our results can help to determine if these new technologies were introduced from outside of the Korean Peninsula, or if they were locally, independently developed.

@@

The migration model argues that the new blade industry, including stemmed points, and the earlier simple flake tool tradition, including large cores, polyhedrals, choppers and handaxes, came from different origins [@bae2010origin; @bae2012nature; @bae2013early]. While the *in situ* model claims that the heterogenic character is the result of indigenous development, the migration model proposes that it is the result of the continuous influx of modern human populations from both north and south. Specifically, the blade technology in claimed to have been introduced from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP, while the simple flake-tool tradition came from southern China [@bae2013early]. The migration of people from the southern route is supported by genetic studies of the Y chromosome, indicating that the O3-M122 M122 haplotype originated from southern East Asia and moved to northeastern Asia, including Korea, at 30-25 ka BP [@shi2005chromosome]. Bae et al. [-@bae2012nature] assume this southern migration could be related to paleoenvironmental fluctuations during the MIS 3 to 2 transition, which made the Yellow Sea/West Sea region open.

Lee [-@lee2013current] argues that the transition to the Korean Late Paleolithic might be more complicated than those models of migration or *in situ* development. He partly agrees with the *in situ* model that simple flake tools had continuously been used in Korea as the result of ancestor-descendant relationships, under conditions of low effective population size. With regard to the blade industry, he claims that low degrees of uniformity and small quantities of blade-associated toolkits indicate an origin outside of Korea, perhaps resulting from trade or migration. We explore these three options, *in situ*, migration, and a mixture of the two, by measuring transmission biases in assemblages of stemmed points.

@@

Cultural transmission theory holds that information about how to behave is acquired through interaction with other individuals and the environment [@boyd1988culture; @richerson1992cultural]. Individuals learn by themselves (e.g. trial and error), or from each other by sharing information. Information can be modified (also known as 'biased') depending on an individual's transmission context and cultural repertoire. Modification of information can occur by recombination, loss, or partial alteration [@eerkens2005cultural; @o2017dual]. Transmission biases can be important loci of changes in material culture, and can be influenced by social contexts of cultural transmission [@heyes1994social; @eerkens2007cultural; @lycett2015cultural; @creanza2017cultural; @kendal2018social]. These biases include guided variation (where individuals learn new behaviors through trial and error), content-based bias (where some aspect of the transmitted instructions, such as cultural preferences, makes them more likely to be adopted), frequency-based bias (where an individual is biased to choose particular behaviors based on their perceived frequency in the population, such as extremely popular or rare behaviors), and indirect bias (where a behavior is transmitted because of its association with other attributes, such as the prestige or skill of other individuals) [@boyd1988culture; @richerson1992cultural; @o2017dual].

Bettinger and Eerkens [-@bettinger1997evolutionary; -@bettinger1999point; -@eerkens2008cultural] have shown how two these biases can be identified in stone artifact assemblages. Their research focused on metric variables of stone points during the introduction of bow and arrow technology in the Great Basin around AD 300-600. They equated guided transmission (where individuals learn new behaviors through trial and error) with high metric variation and low correlation between metric attributes. Conversely, they inferred indirect bias (where a behavior is transmitted because of its association with other attributes, such as the prestige or skill of other individuals) from less variation and more correlated variables. They found that artifacts in the new bow and arrow technology in eastern California have low correlations of basal width and mass, which they interpreted as a result of guided transmission dominating the introduction and spread of these tools. From these results, they inferred a social context in eastern California of distant and unfamiliar neighbors, with limited contact between groups to acquire technology directly. In this context of limited contact, they argue that the new technologies developed largely by trial and error. On the other hand, in central Nevada metric variables are highly correlated, indicating that the bow and arrow was introduced and spread by indirect bias.

Following Bettinger and Eerkens' framework, Garvey [-@garvey2018current] uses simulation to explore the degree of standardization represented by coefficient of variation (CV) values of projectile points from the US Southwest and westernmost southern High Plains. Garvey measured CV values of weight, thickness, width and lengths on two types of projectile points, Washita and Fresno points, from the Henderson site. She then computed simulated CV values according to various three scenarios of three levels of transmission fidelity (i.e. CV = 10%, 5%, and 3%). The observed metrics of the archaeological projectile points are closest to the metrics of the simulated assemblage with 3% CV, which represents "extremely high-fidelity copying". A CV of 3% is the Weber fraction, or threshold of human visual perception, with variation below this value being to subtle for people to notice [@eerkens2000practice]. Garvey's work shows how the cultural transmission of tool-making behaviours can be measured, simulated, and used to interpret the archaeological record.

Due to their focus on archaeological data, Garvey and Bettinger and Eerkens inferred transmission processes from material culture because they were unable to observe them directly. Direct observations have been reported by Mesoudi and O'Brien [-@mesoudi2008cultural] who conducted experimental research using groups of undergraduate students to test the underlying assumptions of Bettinger and Eerkens' study. They simulated model-based bias in projectile point making by providing the design of a model and information of the model's prior success to their research participants. They also simulated guided variation by allowing their participants to explore their own designs. They observed that the majority of participants who were able to choose the previous design copied the most-successful model. Metric attributes of the points made when copied from successful models were more highly correlated than attributes of points made by trial and error. Mesoudi and O'Brien's experimental results are important because they confirm the robustness of previous assumptions about cultural transmission biases, and validate the connection between material culture variation and cultural transmission biases.

@@

Inspired by Bettinger and Eerkens' approach, we use the two contrasting transmission modes, guided variation and indirect bias, to investigate the spread of stemmed point technology during the Korean Late Paleolithic period. We propose a spectrum on which we can locate foraging groups and how they started to make stemmed points, depending on their degree of social isolation or social connectedness. On one end of the spectrum we have socially isolated groups who made stemmed points entirely stimulated by guided variation; and on the other end of the spectrum we have socially connected groups whose knowledge of stemmed points derived from indirect bias on the transmission process.

To contextualize this further, on one end of our spectrum we have socially isolated groups who stayed in physically remote places from other groups, or had unfamiliar neighbors with limited contact between groups. Our assumption for this social context is that individuals or groups acquired the technology of stemmed points by modifying existing flake tool forms through trial-and-error processes to solve problems relating to resource procurement. This trial and error behavior is expected to have left a distinctive signature on metric variables of the stemmed points. If trial and error was the dominant bias in the transmission of knowledge about how to make stemmed points, we expect that the morphological attributes on a stemmed point will be poorly correlated and show high variation within assemblages and between sites.

On the other end of our spectrum, we have socially connected groups that occupied places close to other groups or had regular contact with other groups. Relatively high degrees of social connectivity provide frequent opportunities for observing others and acquiring information. Individuals or groups learned the technology of stemmed points by copying a model from another individual or group. In this scenario, the model, or ideal, stemmed point is likely to have been a highly successful and frequently chosen one, and learners copy all information about the point design as a package. In this social context, indirect bias is the dominant influence on transmission of lithic technology. If indirect bias was prominent in the Korean Paleolithic, we expect that stemmed points should be more standardized, and attributes on the stemmed points will be more correlated, and assemblages from multiple sites will show less variance between and within them.

@@

We used previously developed chronologies that identify three phases in the Korean Late Paleolithic [@seong2015diversity; @Park_2013]. These chronologies were built on radiocarbon ages, and for sites without any, the presence/absence of blades, stemmed point blanks, and toolkit composition. Based on the existence of blade and micro blade and types of stemmed point blanks, we divided the assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blade or flake and the existence of blade in assemblages, and 3) stemmed points made out of blade and the existence of micro blade in assemblages (@tbl-korean-chronology). Applying this division is necessary because some sites such as Yonghodong, Goryeri, Jungmal, and Mungyeong have no radiocarbon dates due to various research limitations. Inferring ages for these sites by analogy to the technological sequences at sites with dates is important for making maximal use of the available archaeological data to increase our sample size. We arranged our `r total_number_of_assemblages` assemblages containing stemmed points in chronological order to facilitate observations of change over time.

@@

To measure the variation among the measurements of the stemmed points, we used Coefficients of Variation (CV) with confidence intervals. The CV is the ratio of the standard deviation to the mean. It has been used as a robust statistical method to calculate the degree of standardization, precision, equality, homogeneity, etc. in various disciplines [@ng2006performance; @panichkitkosolkul2009improved; @panichkitkosolkul2013confidence]. In archaeological studies, CV has been applied to measure the variation between artifacts and test hypotheses about cultural evolutionary processes, including distinguishing between different types of learning biases affecting cultural transmission [@eerkens2001techniques; @eerkens2005cultural; @schillinger2014copying; @garvey2018current]. Eerkens and Bettinger [-@eerkens2001techniques] summarize CV values for artifacts made from various materials and find that the range of CV values for most archaeological datasets is between 2 (i.e. CV for planned production = 1.7) and about 60 (i.e. CV for random production = 57.7) and the values over the upper limit likely reflect intended attempts to differentiate the products. Among the Great Basin Projectiles examined by Eerkens and Bettinger, the average CV was 22 and the range in CV values was 6-55. Considering the lower quality of raw materials in Korea and the median of the CV range (30.5), we selected roughly 25 as the threshold to decide between guided variation and indirect bias. We premised that CV values that are lower or similar to 25 represent guided variation while CV values over 25 reflect indirect bias.

Sample size can impact on accuracy of estimated CV that reflects the corresponding population size. [@toebe2018sample; @kelley2007sample; @vanpool2011quantitative]. In previous archaeological studies, sample sizes for CV values varies from 5 to more than a thousand artifacts [@bettinger1999point; @garvey2018current; @costin1995standardization; @wierer2013variability; @kvamme1996alternative; @rivals2009new]. A variety of methods are available for improving the reliability of CV measurements on small samples. @vanpool2011quantitative proposed a "corrected CV", for a smaller sample (i.e. n\<25). Statistical research has resulted in several methods for computing confidence intervals on CVs to show uncertainty [@koopmans1964confidence; @miller1991asymptomatic; @sharma1994asymptotic; @mckay1932distribution; @vangel1996confidence; @curto2009coefficient; @banik2011estimating; @panichkitkosolkul2013confidence; @mahmoudvand2009two; @gulhar2012comparison]. Although we have a total of `r total_number_of_sp` artifacts, we used the corrected CV formula since we computed CV per assemblages to examine the temporal and regional pattern. We also computed confidence intervals using the Sharma and Krishna model to increase the credibility of the comparisons and be transparent about the precision and accuracy of our results. [@kelley2007sample; @albatineh2014confidence].

To compute correlations between the attributes [@bettinger1999point; @eerkens2008cultural; @garvey2018current], we chose Principal Component Analysis (PCA). PCA is commonly used to detect correlations among multiple variables, and has the advantage of allowing us to investigate the relationship between attributes as well as the impact of each attribute on the overall shape of stemmed points. PCA is often used to reduce dimensionality, detect patterns and clusters in data, indicate relationships between variables as well as cases, and identify which variables most influence these patterns [@cascalheira2018use; @suarez2019life]. PCA is widely used for geometric morphometric analysis of artifacts to extract general patterns of shape variation [@selden2014advances; @suarez2019life; @leplongeon2020backed; @ivanovaite2020all].

All data treatment, analyses and visualization were performed in the R environment [@core2021team]. Our R code and data, including the original artefact images, are fully and openly available online at https://doi.org/10.17605/osf.io/eb8mx to enable full reproducibility [@marwick2017computational].

@@

@fig-plot-cv-all-attributes (panel A) shows a relatively wide distribution of body length (BL) and maximum length (ML) while the other attributes are more similar and shorter (under 2.5 cm). Among these attributes, stem width (SW) is the smallest and most highly standardized, regardless of overall size of the stemmed points. This result implies a similar thickness of a wooden shaft that is attached to the stem.

@@

We used the corrected CV to measure the variations between stemmed points. If the variation is high, the transmission might happen through guided variation while the low variation represents indirect bias. Our results show that CV values for all attributes are distributed from `r min_cv` to `r max_cv` (@fig-plot-cv-all-attributes panel B). Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV is higher. Unlike the distribution of actual attributes, the CVs of body length and maximum length have the lowest values. The lower CV of maximum length indicates they are highly standardized with narrowest confidence intervals. Tang and stem related attributes (e.g. SL, SW, TL, TW) are less standardized with higher CV values and wider confidence intervals than other attributes Among all the attributes, only maximum length has the CV values lower than 25, which could be the result of indirect bias.

@@

To understand the temporal patterns in the modes of cultural transmission, we applied Korea Late Paleolithic chronological phases into our samples to compute CV (@tbl-korean-chronology). Among the three phases of the chronology, we excluded Phase 1 for the analysis because there are only two stemmed points from Yonghodong and Sachang sites to compute. However, this does not mean that there were only two stemmed points for the Phase 1 because we excluded broken pieces from our sample to conduct the morphological analysis.

>>>>

abstract: |
  The onset of the Late Paleolithic period in Korea, represented by the appearance of stemmed points and blades, was a key event in the dispersal of modern humans in Northeast Asia. Previous studies have mainly focussed on possible origin locations of these new technologies. The specific cultural processes of the appearance stemmed points and blades has rarely been considered. We investigate the cultural processes by applying a cultural transmission framework to investigate the social contexts of the emergence of these new technologies. Our main question is: what was the dominant mode of cultural transmission during this time of technological innovation in the Korean Late Paleolithic? Following Bettinger and Eerkens [-@bettinger1999point], we evaluated models of guided variation and indirect bias using data from Korean assemblages containing stemmed points. To evaluate these models and understand the transmission processes, we computed coefficients of variation (CV), and Principal Component Analysis (PCA). We found that information about the new technology was likely transmitted via selective combinations of guided variation and indirect bias. Some attributes including length and width were transmitted through indirect bias, while other attributes appear to have been more due to guided variation. Our results suggest that individuals or groups learned some part of the technology of stemmed points by copying a model from another another individual or group, and then modified the shape according to their specific circumstances.

@@

The application of evolutionary theory to archaeological research has been productive for the investigation of technological transitions and related human behaviors in the remote past [@dunnell1980evolutionary; @lipo1997population; @bettinger1997rediscovering; @bettinger1999point; @mesoudi2008cultural]. For example, archaeologists have used evolutionary theory and methods to study human behavioral ecology, cultural transmission, artifact phylogenetics, and niche construction in the past [@riede2010isn; @o2017dual; @garvey2018current]. In this paper we used cultural transmission theory to investigate technological transitions during the Korean Late Paleolithic. The primary technological innovation is during this time was the appearance of new tools such as stemmed points and blades. The appearance of these new lithic technologies may represent the first arrival of modern humans in this region, and so they may also represent a key event related to modern human dispersal throughout East Asia [@seong2009emergence]. Previous studies on the origin of stemmed points have mainly focused on their possible origin locations, connecting Korea with global patterns of modern human dispersal. However, questions about the specific cultural processes and social contexts of this technological change in Korea remain largely unanswered [@seong2008tanged; @bae2010origin; @bae2017origin]. In this research, we apply a cultural transmission framework to investigate the social contexts of the emergence of new technologies in the Korean Late Paleolithic. Our main question is: what was the dominant mode of cultural transmission for technological innovation in the Korean Late Paleolithic? Additionally we ask: Do the modes of cultural transmission vary over time and space? We consider three possible modes of cultural transmission: guided variation (trial and error), indirect bias (copying others), or a combination of the two. Our results have implications for determining if these new technologies were introduced from outside of the Korean Peninsula, or if they were locally, independently developed.

@@

In contrast, the migration model argues that the new blade industry, including stemmed points, and the earlier simple flake tool tradition, including large cores, polyhedrals, choppers and handaxes, came from different origins [@bae2010origin; @bae2012nature; @bae2013early]. While the *in situ* model claims that the heterogenic character is the result of indigenous development, the migration model proposes that it is the result of the continuous influx of modern human populations from both north and south. Specifically, the blade technology in claimed to have been introduced from Siberia, Mongolia, or other regions of northeast China following the Liaohe and Sunghe rivers around 35 ka BP, while the simple flake-tool tradition came from southern China [@bae2013early]. The migration of people from the southern route is supported by genetic studies of the Y chromosome, indicating that the O3-M122 M122 haplotype originated from southern East Asia and moved to northeastern Asia, including Korea, at 30-25 ka BP [@shi2005chromosome]. Bae et al. [-@bae2012nature] assume this southern migration could be related to paleoenvironmental fluctuations during the MIS 3 to 2 transition, which made the Yellow Sea/West Sea region open.

Lee [-@lee2013current] argues that the transition to the Korean Late Paleolithic might be more complicated than either models of migration or *in situ* development. He partly agrees with the *in situ* model that simple flake tools had continuously been used in Korea as the result of ancestor-descendant relationships, under conditions of low effective population size. With regard to the blade industry, he claims that low degrees of uniformity and small quantities of blade-associated toolkits indicate an origin outside of Korea, perhaps resulting from trade or migration. We explore these three options, *in situ*, migration, and a mixture of the two, by measuring transmission biases in assemblages of stemmed points.

@@

Cultural transmission theory holds that information about how to behave is acquired through interaction with other individuals and the environment [@boyd1988culture; @richerson1992cultural]. Individuals learn by themselves (e.g. trial and error), or from each other by sharing information. Information can be modified (also known as 'biased') depending on an individual's transmission context and cultural repertoire. Modification of information can occur by recombination, loss, or partial alteration [@eerkens2005cultural; @o2017dual]. Transmission biases can be important loci of changes in material culture, and can be influenced by the social contexts of cultural transmission [@heyes1994social; @eerkens2007cultural; @lycett2015cultural; @creanza2017cultural; @kendal2018social]. These biases include guided variation (where individuals learn new behaviors through trial and error), content-based bias (where some aspect of the transmitted instructions, such as cultural preferences, makes them more likely to be adopted), frequency-based bias (where an individual is biased to choose particular behaviors based on their perceived frequency in the population, such as extremely popular or rare behaviors), and indirect bias (where a behavior is transmitted because of its association with other attributes, such as the prestige or skill of other individuals) [@boyd1988culture; @richerson1992cultural; @o2017dual].

Bettinger and Eerkens [-@bettinger1997evolutionary; -@bettinger1999point; -@eerkens2008cultural] have shown how two of these biases can be identified in stone artifact assemblages. Their research focused on metric variables of stone points during the introduction of bow and arrow technology in the Great Basin around AD 300-600. They equated guided transmission (where individuals learn new behaviors through trial and error) with high metric variation and low correlation between metric attributes. Conversely, they inferred indirect bias (where a behavior is transmitted because of its association with other attributes, such as the prestige or skill of other individuals) from less variation and more correlated variables. They found that artifacts in the new bow and arrow technology in eastern California have low correlations of basal width and mass, which they interpreted as a result of guided transmission dominating the introduction and spread of these tools. From these results, they inferred a social context in eastern California of distant and unfamiliar neighbors, with limited contact between groups to acquire technology directly. In this context of limited contact, they argue that the new technologies developed largely by trial and error. On the other hand, in central Nevada they found that metric variables are highly correlated, indicating that the bow and arrow was introduced and spread by indirect bias.

Following Bettinger and Eerkens' framework, Garvey [-@garvey2018current] uses simulation to explore the degree of standardization represented by coefficient of variation (CV) values of projectile points from the US Southwest and westernmost southern High Plains. Garvey measured CV values of weight, thickness, width and lengths on two types of projectile points, Washita and Fresno points, from the Henderson site. She then computed simulated CV values according to three scenarios of different levels of transmission fidelity (i.e. CV = 10%, 5%, and 3%). The observed metrics of the archaeological projectile points are closest to the metrics of the simulated assemblage with 3% CV, which represents "extremely high-fidelity copying". A CV of 3% is the Weber fraction, or threshold of human visual perception, with variation below this value being too subtle for people to notice [@eerkens2000practice]. Garvey's work shows how the cultural transmission of tool-making behaviours can be measured, simulated, and used to interpret the archaeological record.

Garvey and Bettinger and Eerkens inferred transmission processes from material culture because they were unable to observe them directly. Direct observations have been reported by Mesoudi and O'Brien [-@mesoudi2008cultural] who conducted experimental research using groups of undergraduate students to test the underlying assumptions of Bettinger and Eerkens' study. They simulated model-based bias in projectile point making by providing the design of a model and information of the model's prior success to their research participants. They also simulated guided variation by allowing their participants to explore their own designs. They observed that the majority of participants who were able to choose the previous design copied the most-successful model. Metric attributes of the points made when copied from successful models were more highly correlated than attributes of points made by trial and error. Mesoudi and O'Brien's experimental results are important because they confirm the robustness of previous assumptions about cultural transmission biases, and validate the connection between material culture variation and cultural transmission biases.

@@

Inspired by Bettinger and Eerkens' approach, we use two contrasting transmission modes, guided variation and indirect bias, to investigate the spread of stemmed point technology during the Korean Late Paleolithic period. We propose a spectrum on which we can locate foraging groups and how they started to make stemmed points, depending on their degree of social isolation or social connectedness. On one end of the spectrum we have socially isolated groups who made stemmed points entirely stimulated by guided variation; and on the other end of the spectrum we have socially connected groups whose knowledge of stemmed points derived from indirect bias on the transmission process.

To contextualize this further, on one end of our spectrum we have socially isolated groups who stayed in physically remote places from other groups, or had unfamiliar neighbors with limited contact between groups. Our assumption for this social context is that individuals or groups acquired the technology of stemmed points by modifying existing flake tool forms through trial-and-error processes to solve problems relating to resource procurement (cf. Seong's *in situ* model). This trial and error behavior is expected to have left a distinctive signature on metric variables of the stemmed points. If trial and error was the dominant bias in the transmission of knowledge about how to make stemmed points, we expect that the morphological attributes on a stemmed point will be poorly correlated and show high variation within assemblages and between sites.

On the other end of our spectrum, we have socially connected groups that occupied places close to other groups or had regular contact with other groups. Relatively high degrees of social connectivity provide frequent opportunities for observing others and acquiring information. Individuals or groups learned the technology of stemmed points by copying a model from another individual or group. In this scenario, the model, or ideal, stemmed point is likely to have been a highly successful and frequently chosen one, and learners copy all information about the point design as a package (cf. Bae's migration model). In this social context, indirect bias is the dominant influence on transmission of lithic technology. If indirect bias was prominent in the Korean Paleolithic, we expect that stemmed points should be more standardized, and attributes on the stemmed points will be more correlated, and assemblages from multiple sites will show less variance between and within them.

@@

To analyse change over time, we used previously developed chronologies that identify three phases in the Korean Late Paleolithic [@seong2015diversity; @Park_2013]. These chronologies were built on radiocarbon ages, and for sites without any, the presence/absence of blades, stemmed point blanks, and toolkit composition. Based on the existence of blade and micro blade and types of stemmed point blanks, we divided the assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blade or flake and the existence of blade in assemblages, and 3) stemmed points made out of blade and the existence of micro blade in assemblages (@tbl-korean-chronology). Applying this division is necessary because some sites such as Yonghodong, Goryeri, Jungmal, and Mungyeong have no radiocarbon dates due to various research limitations. Inferring ages for these sites by analogy to the technological sequences at sites with dates is important for making maximal use of the available archaeological data to increase our sample size. We arranged our `r total_number_of_assemblages` assemblages containing stemmed points in chronological order to analyse them using these three phases to facilitate observations of change over time.

@@

To measure the variation among the measurements of the stemmed points, we computed Coefficients of Variation (CV) with confidence intervals. The CV is the ratio of the standard deviation to the mean. It has been used as a robust statistical method to calculate the degree of standardization, precision, equality, homogeneity, etc. in various disciplines [@ng2006performance; @panichkitkosolkul2009improved; @panichkitkosolkul2013confidence]. In archaeological studies, CV has been applied to measure the variation between artifacts and to test hypotheses about cultural evolutionary processes, including distinguishing between different types of learning biases affecting cultural transmission [@eerkens2001techniques; @eerkens2005cultural; @schillinger2014copying; @garvey2018current]. Eerkens and Bettinger [-@eerkens2001techniques] summarized CV values for artifacts made from various materials and find that the range of CV values for most archaeological datasets is between 2 (i.e. CV for planned production = 1.7) and about 60 (i.e. CV for random production = 57.7) and the values over the upper limit likely reflect  attempts to intentionally differentiate the products. Among the Great Basin projectiles examined by Eerkens and Bettinger, the average CV was 22 and the range in CV values was 6-55. Considering the lower quality of raw materials in Korea compared to the Great Basin, and the median of the CV range (30.5) for the Korean lithics, we selected 25 as the threshold to decide between guided variation and indirect bias. We premised that CV values that are lower than 25 represent guided variation while CV values over 25 reflect indirect bias.

Sample size can impact the accuracy of CV estimates. [@toebe2018sample; @kelley2007sample; @vanpool2011quantitative]. In previous archaeological studies, sample sizes for CV values has varied from five to more than a thousand artifacts [@bettinger1999point; @garvey2018current; @costin1995standardization; @wierer2013variability; @kvamme1996alternative; @rivals2009new]. A variety of methods are available for improving the reliability of CV measurements when using small samples. For example, @vanpool2011quantitative proposed a "corrected CV", for a smaller sample (i.e. n\<25). Statistical research has resulted in several methods for computing confidence intervals on CVs to show uncertainty [@koopmans1964confidence; @miller1991asymptomatic; @sharma1994asymptotic; @mckay1932distribution; @vangel1996confidence; @curto2009coefficient; @banik2011estimating; @panichkitkosolkul2013confidence; @mahmoudvand2009two; @gulhar2012comparison]. Although we have a total of `r total_number_of_sp` artifacts, most of our analyses here involves comparisons of smaller subsets of this sample. For example to explore temporal and regional patterns we compared samples of well under 25 pieces. Given these relatively small sample sizes in our study, we used the corrected CV formula, which we have implemented here in a function in the R programming language for others to use. We also computed confidence intervals using method proposed by @sharma1994asymptotic to increase the credibility of the comparisons and be transparent about the precision and accuracy of our results [@kelley2007sample; @albatineh2014confidence].

To compute correlations between the attributes [@bettinger1999point; @eerkens2008cultural; @garvey2018current], we used Principal Component Analysis (PCA). PCA is commonly used to detect correlations among multiple variables, and has the advantage of allowing us to investigate the relationship between attributes as well as the impact of each attribute on the overall shape of stemmed points. PCA is often used to reduce dimensionality, detect patterns and clusters in data, indicate relationships between variables as well as cases, and identify which variables most influence these patterns [@cascalheira2018use; @suarez2019life]. PCA is widely used for geometric morphometric analysis of artifacts to extract general patterns of shape variation [@selden2014advances; @suarez2019life; @leplongeon2020backed; @ivanovaite2020all].

All data preparation, analyses and visualization were performed in the R environment [@core2021team]. Our R code and data, including the original artefact images, are fully and openly available in our compendium [@Marwick2018compendium] online at https://doi.org/10.17605/osf.io/eb8mx to enable transparency and reproducibility [@marwick2017computational].

@@

@@

@fig-plot-cv-all-attributes (panel A) shows a relatively wide distribution of body length (BL) and maximum length (ML) while the other attributes are more similar and shorter (under 2.5 cm). Among these attributes, stem width (SW) is the smallest, regardless of overall size of the stemmed points. This result may imply that the stem width was the most constrained dimension of the artefact. We might speculate that this is due to the requirements of halfting with wooden shafts, where the shafts were selected from a small range of diameters that optimised strength, weight and flexibility.

Our results show that CV values for all attributes are distributed from `r min_cv` to `r max_cv` (@fig-plot-cv-all-attributes). Compared to the CVs of Great Basin projectile points [@eerkens2001techniques], the average CV for the Korean artefacts is higher. Unlike the distribution of the linear measurements, the CVs of body length and maximum length have the lowest values. The lower CV of maximum length indicates they are highly standardized with narrowest confidence intervals. Tang and stem related attributes (e.g. SL, SW, TL, TW) are less standardized with higher CV values and wider confidence intervals than other attributes Among all the attributes, only maximum length has the CV values lower than 25, which suggests the transmission of this attribute was influenced by indirect bias.

@@

To understand the temporal patterns in the modes of cultural transmission, we grouped our assemblages into the three Korean Late Paleolithic chronological phases and computed CV values for each phase (@tbl-korean-chronology). Among the three phases of the chronology, we excluded Phase 1 for the analysis because there are only two stemmed points from Yonghodong and Sachang sites to compute. However, this does not mean that there were only two stemmed points for the Phase 1 because we excluded broken pieces from our sample to conduct the morphological analysis.
